Live Programming with Typed Holes

### 2018-01-11 calculus.tex
adding rule names, cleaning up after talking to cyrus
missing? FHCastFinal rule
removing icast rule
changes?

\begin{figure}[!ht]
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{HTyp} & \htau & ::= &
  b ~\vert~
  \tarr{\htau}{\htau} ~\vert~
  % \tprod{\htau}{\htau} ~\vert~
  % \tsum{\htau}{\htau} ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  c ~\vert~
  x ~\vert~
  \halam{x}{\htau}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  % \hpair{\hexp}{\hexp} ~\vert~
  % \hprj{i}{\hexp} ~\vert~
  % \hinj{i}{\hexp} ~\vert~
  % \hcase{\hexp}{x}{\hexp}{x}{\hexp} ~\vert~
  % \hadd{\hexp}{\hexp} ~\vert~
  \hehole{u} ~\vert~
  \hhole{\hexp}{u} ~\vert~
  {\hlam{x}{\hexp}} ~\vert~
  \hexp : \htau\\
\mathsf{Mark} & \markname{} & ::= &
  \evaled{} ~\vert~  \unevaled{}\\
 \mathsf{DHExp} & \dexp  & ::= &
  c ~\vert~
  x ~\vert~
  {\halam{x}{\htau}{\dexp}} ~\vert~
  \hap{\dexp}{\dexp} ~\vert~
  % \hpair{\dexp}{\dexp} ~\vert~
  % \hprj{i}{\dexp} ~\vert~
  % \hinj{i}{\dexp} ~\vert~
  % \hcase{\dexp}{x}{\dexp}{x}{\dexp} ~\vert~
  % \hadd{\dexp}{\dexp} ~\vert~
  \dehole{\mvar}{\subst}{\markname} ~\vert~
  \dhole{\dexp}{\mvar}{\subst}{\markname} ~\vert~
  \dcast{\htau}{\dexp}\\
\end{array}$
\caption{Syntax of H-types, H-expressions and dynamic H-expressions. We write $x$ to range over variables, $u$ over hole names and $\sigma$ over finite substitutions.}
\label{fig:HTyp}
\label{fig:HExp}
\end{figure}

@@

\inferrule[TAEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{\markname}}{\htau}
}

\inferrule[TANEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\\\
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{\markname}}{\htau}
}

@@

\begin{figure}[!ht]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcll}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y & (y \neq x)\\
{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
&=&
{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
& (y \neq x)
\\
{[\dexp_1 / x] \dap{d_2}{d_3}}
&=&
{\dapP{[\dexp_1 / x] d_2}{[\dexp_1 / x] d_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}{m}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}{m}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{m}}
&=&
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{m}
\\
{[\dexp_1 / x] \dcast{\htau}{\dexp}}
&=&
\dcast{\htau}{[\dexp_1 / x] \dexp}
\end{array}
\]
\end{figure}

@@

\inferrule[ENEHole]{
  \isErr{\Delta}{\dexp}
}{
  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{m}}
}

@@

\begin{figure}
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{EvalCtx} & \evalctx & ::= &
  \evalhole ~\vert~
  \halam{x}{\htau}{\evalctx} ~\vert~
  \hap{\evalctx}{\dexp} ~\vert~
  \hap{\dexp}{\evalctx} ~\vert~
  \dhole{\evalctx}{\mvar}{\subst}{\markname} ~\vert~
  \dcast{\htau}{\evalctx}
\end{array}$

@@

\inferrule[ECLam]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\halam{x}{\htau}{\evalctx}}
}

@@

\inferrule[ECNEHole]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{\markname}}
}

@@

\judgbox{\selectEvalCtx{d}{\evalctx}{d'}}{$d$ is the result of filling the
hole in $\evalctx$ with $d'$}
\begin{mathpar}
\inferrule[FDot]{ }{
  \selectEvalCtx{d}{\evalhole}{d}
}

%% \todo{d used to be c -- TODO IEV}

\inferrule[FLam]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
}

\inferrule[FAp1]{
  \isFinal{d_1}\\
  \selectEvalCtx{d_2}{\evalctx}{d_2'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{d_1}{\evalctx}}{d_2'}
}

\inferrule[FAp2]{
  \selectEvalCtx{d_1}{\evalctx}{d_1'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{\evalctx}{d_2}}{d_1'}
}

%% I think these rules are not needed, if you make FDot work for any dhexp
%% d. The only reason they might be needed is because they're specified to
%% have check marks, not x marks, but i don't think that's important. i
%% think the more general FDot makes sense. -- TODO IEV
%%
%%
%% \inferrule[\todo{??}]{ }{
%%   \selectEvalCtx{\dehole{\mvar}{\subst}{\evaled}}{\evalhole}{\dehole{\mvar}{\subst}{\evaled}}
%% }

%% \inferrule[\todo{??}]{ }{
%%   \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
%% }

\inferrule[\todo{??}]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{\evalctx}{\mvar}{\subst}{\unevaled}}{d'}
}

\inferrule[\todo{??}]{
  \isFinal{d}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\unevaled}}
}

\inferrule[FCast]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
}
\end{mathpar}

@@

\begin{theorem}[Progress]
If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
$\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or
$\stepsToD{\Delta}{\dexp}{\dexp'}$ for some $\dexp'$.
\end{theorem}

>>>>

\begin{figure}[!ht]
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{HTyp} & \htau & ::= &
  b ~\vert~
  \tarr{\htau}{\htau} ~\vert~
  % \tprod{\htau}{\htau} ~\vert~
  % \tsum{\htau}{\htau} ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  c ~\vert~
  x ~\vert~
  \halam{x}{\htau}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  % \hpair{\hexp}{\hexp} ~\vert~
  % \hprj{i}{\hexp} ~\vert~
  % \hinj{i}{\hexp} ~\vert~
  % \hcase{\hexp}{x}{\hexp}{x}{\hexp} ~\vert~
  % \hadd{\hexp}{\hexp} ~\vert~
  \hehole{u} ~\vert~
  \hhole{\hexp}{u} ~\vert~
  {\hlam{x}{\hexp}} ~\vert~
  \hexp : \htau\\
% \mathsf{Mark} & \markname{} & ::= &
%   \evaled{} ~\vert~  \unevaled{}\\
 \mathsf{DHExp} & \dexp  & ::= &
  c ~\vert~
  x ~\vert~
  {\halam{x}{\htau}{\dexp}} ~\vert~
  \hap{\dexp}{\dexp} ~\vert~
  % \hpair{\dexp}{\dexp} ~\vert~
  % \hprj{i}{\dexp} ~\vert~
  % \hinj{i}{\dexp} ~\vert~
  % \hcase{\dexp}{x}{\dexp}{x}{\dexp} ~\vert~
  % \hadd{\dexp}{\dexp} ~\vert~
  \dehole{\mvar}{\subst}{} ~\vert~
  \dhole{\dexp}{\mvar}{\subst}{} ~\vert~
  \dcast{\htau}{\dexp}\\
\end{array}$
\caption{Syntax of H-types, H-expressions and dynamic H-expressions. We write $x$ to range over variables, $u$ over hole names and $\sigma$ over finite substitutions.}
\label{fig:HTyp}
\label{fig:HExp}
\end{figure}

@@

\inferrule[TAEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dehole{u}{\sigma}{}}{\htau}
}

\inferrule[TANEHole]{
  \Dbinding{u}{\hGamma'}{\htau} \in \Delta\\\\
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta}{\hGamma}{\dhole{\dexp}{u}{\sigma}{}}{\htau}
}

@@

\begin{figure}[!ht]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcll}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y & (y \neq x)\\
{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
&=&
{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
& (y \neq x)
\\
{[\dexp_1 / x] \dap{d_2}{d_3}}
&=&
{\dapP{[\dexp_1 / x] d_2}{[\dexp_1 / x] d_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}{}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}{}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}{}}
&=&
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}{}
\\
{[\dexp_1 / x] \dcast{\htau}{\dexp}}
&=&
\dcast{\htau}{[\dexp_1 / x] \dexp}
\end{array}
\]
\end{figure}

@@

\inferrule[ENEHole]{
  \isErr{\Delta}{\dexp}
}{
  \isErr{\Delta}{\dhole{\dexp}{\mvar}{\subst}{}}
}

@@

%%   \halam{x}{\htau}{\evalctx} ~\vert~

\begin{figure}
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{EvalCtx} & \evalctx & ::= &
  \evalhole ~\vert~
  \hap{\evalctx}{\dexp} ~\vert~
  \hap{\dexp}{\evalctx} ~\vert~
  \dhole{\evalctx}{\mvar}{\subst}{} ~\vert~
  \dcast{\htau}{\evalctx}
\end{array}$

@@

%% \inferrule[ECLam]{
%%   \isevalctx{\evalctx}
%% }{
%%   \isevalctx{\halam{x}{\htau}{\evalctx}}
%% }

@@

\inferrule[ECNEHole]{
  \isevalctx{\evalctx}
}{
  \isevalctx{\dhole{\evalctx}{\mvar}{\subst}{}}
}

@@

\judgbox{\selectEvalCtx{d}{\evalctx}{d'}}{$d$ is the result of filling the
hole in $\evalctx$ with $d'$}
\begin{mathpar}
\inferrule[FHConst]{ }{
  \selectEvalCtx{c}{\evalhole}{c}
}

%% \inferrule[FLam]{
%%   \selectEvalCtx{d}{\evalctx}{d'}
%% }{
%%   \selectEvalCtx{\halam{x}{\htau}{d}}{\halam{x}{\htau}{\evalctx}}{d'}
%% }

\inferrule[FHAp1]{
  \isFinal{d_1}\\
  \selectEvalCtx{d_2}{\evalctx}{d_2'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{d_1}{\evalctx}}{d_2'}
}

\inferrule[FHAp2]{
  \selectEvalCtx{d_1}{\evalctx}{d_1'}
}{
  \selectEvalCtx{\hap{d_1}{d_2}}{\hap{\evalctx}{d_2}}{d_1'}
}

\inferrule[FHEHole]{ }{
  \selectEvalCtx{\dehole{\mvar}{\subst}{}}{\evalhole}{\dehole{\mvar}{\subst}{}}
}

\inferrule[FHNEHoleEvaled]{ }{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\evaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\evaled}}
}

\inferrule[FHNEHoleInside]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\dhole{\evalctx}{\mvar}{\subst}{\unevaled}}{d'}
}

\inferrule[FHNEHoleFinal]{
  \isFinal{d}
}{
  \selectEvalCtx{\dhole{d}{\mvar}{\subst}{\unevaled}}{\evalhole}{\dhole{d}{\mvar}{\subst}{\unevaled}}
}

\inferrule[FHCast]{
  \selectEvalCtx{d}{\evalctx}{d'}
}{
  \selectEvalCtx{\dcast{\htau}{d}}{\dcast{\htau}{\evalctx}}{d'}
}

\inferrule[FHCastFinal]{
  \isFinal{d}
}{
  \selectEvalCtx{\dcast{\htau}{d}}{\evalhole}{\dcast{\htau}{d}}
}
\end{mathpar}

@@

\begin{theorem}[Progress]
If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\stepsToD{\Delta}{\dexp}{\dexp'}$ or 
$\isValue{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\isIndet{\dexp}$. 
% for some $\dexp'$.
\end{theorem}
